Enforced Invariant Confluence

### 2020-01-14 doc/the_vldb_journal_paper/sections/invariant_closure.tex
Added proofs to invariant closure section.

\thmref{IclosureImpliesIconfluence} states that \invariantclosure{} is
sufficient for \invariantconfluence{}. Intuitively, recall that our system model ensures
that transaction execution preserves the invariant, so if merging states also
preserves the invariant and if our start state satisfies the invariant, then
inductively it is impossible for us to reach a state that doesn't satisfy the
invariant.

This is good news because checking if an object is \invariantclosed{} is more
straightforward than checking if it is \invariantconfluent{}. Existing systems
typically use an SMT solver like Z3 to check if an object is
\invariantclosed{}~\cite{de2008z3, balegas2015putting, gotsman2016cause}. If it
is, then by \thmref{IclosureImpliesIconfluence}, it is \invariantconfluent{}.
Unfortunately, \invariantclosure{} is \emph{not} necessary for
\invariantconfluence{}, so if an object is \emph{not} \invariantclosed{}, these
systems cannot conclude that the object is \emph{not} \invariantconfluent{}.
The reason why \invariantclosure{} is not necessary for \invariantconfluence{}
is best explained through an example.

@@

\begin{example}\examplelabel{Z2}
  Let $O = (\ints \times \ints, \join)$ consist of pairs $(x, y)$ of integers
  where $(x_1, y_1) \join (x_2, y_2) = (\max(x_1, x_2), \max(y_1, y_2))$. Our
  start state $s_0 \in \ints \times \ints$ is the point $(0, 0)$. Our set $T$
  of transactions consists of two transactions: $t_{x+1}((x, y)) = (x + 1, y)$
  which increments $x$ and $t_{y-1}((x, y)) = (x, y - 1)$ which decrements $y$.
  Our invariant $I = \setst{(x, y) \in \ints \times \ints}{xy \leq 0}$ consists
  of all points $(x, y)$ where the product of $x$ and $y$ is non-positive.

  The invariant and the set of reachable states are illustrated in \figref{Z2}
  in which we draw each state $(x, y)$ as a point in space. The invariant
  consists of the second and fourth quadrant, while the reachable states consist only
  of the fourth quadrant. From this, it is immediate that the reachable states
  are a subset of the invariant, so $O$ is \invariantconfluent{}. However,
  letting $s_1 = (-1, 1)$ and $s_2 = (1, -1)$, we see that $O$ is not
  \invariantclosed{}. $I(s_1)$ and $I(s_2)$, but letting $s_3 = s_1 \join s_2 =
  (1, 1)$, we see $\lnot I(s_3)$.
\end{example}

@@

\begin{theorem}\thmlabel{IclosureEquivalentIconfluence}
  Consider an object $O = (S, \join)$, a start state $s_0 \in S$, a set of
  transactions $T$, and an invariant $I$. If the invariant is a subset of the
  reachable states (i.e. $I \subseteq \setst{s \in
  S}{\sTIreachablepredicate{s}}$), then
  \[
    (\text{$I(s_0)$ and $O$ is \Iclosed{}}) \iff \text{$O$ is \sTIconfluent{}}
  \]
\end{theorem}
\begin{proof}
  The forward direction of \thmref{IclosureEquivalentIconfluence} follows
  immediately from \thmref{IclosureImpliesIconfluence}. The backward direction
  holds because any two invariant satisfying states $s_1$ and $s_2$ must be
  reachable (by assumption), so their join $s_1 \join s_2$ is also reachable.
  And because $O$ is \sTIconfluent{}, all reachable points, including $s_1
  \join s_2$, satisfy the invariant.
\end{proof}

>>>>

\begin{proof}
\thmref{IclosureImpliesIconfluence} states that \invariantclosure{} is
sufficient for \invariantconfluence{}. To prove this informally, recall that
our system model ensures that transaction execution preserves the invariant, so
if merging states also preserves the invariant and if our start state satisfies
the invariant, then inductively it is impossible for us to reach a state that
doesn't satisfy the invariant.

More formally, consider an arbitrary state $s$ that is reachable via expression
$e$. We prove by structural induction that for every expression $e$, if $e$ is
\sTIreachable{} then $I(e)$.
\begin{itemize}
  \item \textbf{Case $s_0$.}
    $s_0$ is reachable and satisfies the invariant by assumption.

  \item \textbf{Case $t(e)$.}
    If $t(e)$ is reachable, then it satisfies the invariant by definition.

  \item \textbf{Case $e_1 \join e_2$.}
    Let $e_1 \join e_2$ be \sTIreachable{}. Then, $\sTIreachablepredicate{e_1}$
    and $\sTIreachablepredicate{e_2}$. By the inductive hypothesis, $I(e_1)$
    and $I(e_2)$. By \invariantclosure{}, $I(e_1 \join e_2)$.
\end{itemize}
\end{proof}

\thmref{IclosureImpliesIconfluence} is good news because checking if an object
is \invariantclosed{} is more straightforward than checking if it is
\invariantconfluent{}. Existing systems typically use an SMT solver like Z3 to
check if an object is \invariantclosed{}~\cite{de2008z3, balegas2015putting,
gotsman2016cause}. If it is, then by \thmref{IclosureImpliesIconfluence}, it is
\invariantconfluent{}.  Unfortunately, \invariantclosure{} is \emph{not}
necessary for \invariantconfluence{}, so if an object is \emph{not}
\invariantclosed{}, these systems cannot conclude that the object is \emph{not}
\invariantconfluent{}. The reason why \invariantclosure{} is not necessary for
\invariantconfluence{} is best explained through an example.

@@

\begin{example}\examplelabel{Z2}
  Let $O = (\ints \times \ints, \join)$ consist of pairs $(x, y)$ of integers
  where
  \[
    \hfill (x_1, y_1) \join (x_2, y_2) = (\max(x_1, x_2), \max(y_1, y_2)) \hfill
  \]
  Our start state $s_0 \in \ints \times \ints$ is the point $(0, 0)$. Our set
  $T$ of transactions consists of two transactions: $t_{x+1}((x, y)) = (x + 1,
  y)$ which increments $x$ and $t_{y-1}((x, y)) = (x, y - 1)$ which decrements
  $y$.  Our invariant $I = \setst{(x, y) \in \ints \times \ints}{xy \leq 0}$
  consists of all points $(x, y)$ where the product of $x$ and $y$ is
  non-positive.

  The invariant and the set of reachable states are illustrated in \figref{Z2}
  in which we draw each state $(x, y)$ as a point in $\ints^2$. The invariant
  consists of the second and fourth quadrant, while the reachable states consist only
  of the fourth quadrant. From this, it is immediate that the reachable states
  are a subset of the invariant, so $O$ is \invariantconfluent{}. However,
  letting $s_1 = (-1, 1)$ and $s_2 = (1, -1)$, we see that $O$ is not
  \invariantclosed{}. $I(s_1)$ and $I(s_2)$, but letting $s_3 = s_1 \join s_2 =
  (1, 1)$, we see $\lnot I(s_3)$.
\end{example}

@@

\begin{theorem}\thmlabel{IclosureEquivalentIconfluence}
  Consider an object $O = (S, \join)$, a start state $s_0 \in S$, a set of
  transactions $T$, and an invariant $I$. If the invariant is a subset of the
  reachable states (i.e. $I \subseteq \setst{s \in
  S}{\sTIreachablepredicate{s}}$), then
  \[
    (\text{$I(s_0)$ and $O$ is \Iclosed{}}) \iff \text{$O$ is \sTIconfluent{}}
  \]
\end{theorem}

\begin{proof}
  The forward direction of \thmref{IclosureEquivalentIconfluence} follows
  immediately from \thmref{IclosureImpliesIconfluence}. The backward direction
  holds because any two invariant satisfying states $s_1$ and $s_2$ must be
  reachable (by assumption), so their join $s_1 \join s_2$ is also reachable.
  And because $O$ is \sTIconfluent{}, all reachable points, including $s_1
  \join s_2$, satisfy the invariant. Moreover $s_0$ is reachable, so $I(s_0)$
  because $O$ is \sTIconfluent{}.
\end{proof}

### 2020-01-14 doc/the_vldb_journal_paper/sections/invariant_confluence.tex
Added proofs to invariant closure section.

\subsection{Equivalence to Existing Definition}
Our definition of invariant confluence is different than the original
definition given in~\cite{bailis2014coordination}, but the difference is merely
cosmetic. We now prove that the two definitions are equivalent.

>>>>

\subsection{Equivalence to Existing Definition}
Our definition of \invariantconfluence{} is different than the original
definition given in~\cite{bailis2014coordination}, but the difference is merely
cosmetic. We now prove that the two definitions are equivalent.
